Integrated nested Laplace approximations for extended latent Gaussian models

### 2022-11-17 src/docs_paper/paper.Rmd
Remove typo
Add backwards version of the algorithm too (as in defining objects mathematically, rather than how you would code it)

# Fast inference methods\label{sec:fastinferencemethods}

@@

## Algorithm\label{sec:algorithm}

Given a \textsf{C++} user template `model.cpp` for the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adative Gauss-Hermite quadrature.

1. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(\btheta, \y)$
$$
\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}
\end{align*}

@@

3. Obtain an unnormalised nested approximation to the posterior marginal of the $i$th latent effect by
$$
\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) , \y) \omega(\z).
$$
The nodes and weights $\{\mathcal{Q}(m, k), \omega\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{AQ}(x_i \, | \, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following Laplace approximation
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x_{-i}(x_i, \btheta), \mathbf{H}_{-i, -i}(x_i, \btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
\end{align*}

4. Normalise $\tilde p_\text{LA}(x_i, \y)$ using $\tilde p_{\text{AQ}}(\y)$ to obtain
$$
\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
$$

>>>>

# Fast approximate inference method\label{sec:fastinferencemethods}

@@

## Forwards algorithm\label{sec:algorithm}

Given a \textsf{C++} user template `model.cpp` for the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adaptive Gauss-Hermite quadrature.

1. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(\btheta, \y)$
$$
\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}

@@

3. Obtain an unnormalised nested approximation to the posterior marginal of the $i$th latent effect by
$$
\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
$$
The nodes and weights $\{\mathcal{Q}(m, k), \omega\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{LA}(x_i, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following Laplace approximation
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x_{-i}(x_i, \btheta), \mathbf{H}_{-i, -i}(x_i, \btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
\end{align*}
Optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$.

4. Normalise $\tilde p_\text{LA}(x_i, \y)$ using $\tilde p_{\text{AQ}}(\y)$ to obtain
$$
\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
$$
which may be evaluated for some choice of values $x_i \in \{\ldots\}$.

## Backward algorithm\label{sec:algorithm}

1. Calculate
\begin{align*}
\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
\end{align*}
where
$$
\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
2. Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $m = \dim(\btheta)$, $k$ nodes per dimension, as follows
\begin{align*}
\btheta(\z) &= \hat{\btheta} + \mathbf{L} \z, \\
\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top.
\end{align*}
We typically set $k = 3$ such that there are $3^m$ nodes in total.
3. Use this quadrature rule to calculate $\tilde p_{\text{AQ}}(\y)$ as follows
$$
\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z).
$$
4. For $x_i \in \{\ldots\}$ and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
\end{align*}
where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$.

5. For each $x_i$ calculate
$$
\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
$$
where
$$
\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
$$
and
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
$$
